# Data Wrangling {#sec-wrangle}

## Intended Learning Outcomes {#sec-ilo-wrangle .unnumbered}

* Be able to select and filter data for relevance
* Be able to create new columns and edit existing ones
* Be able to handle missing data

## Functions used {#functions-wrangle}

```{r, include = FALSE}
# load tidyverse packages separately so auto-links work in `func()` notation
library(readr)
library(dplyr)
library(tidyr)
library(tibble)
library(janitor)
```

* built-in (you can always use these without loading any packages)
    * base:: `round()`, `mean()`, `min()`, `max()`, ``sum()`, `as.character()`, `as.numeric()`
    * utils:: `head()`, `packageVersion()`
* tidyverse (you can use all these with `library(tidyverse)`)
    * readr:: `read_csv()`
    * dplyr:: `filter()`, `arrange()`, `mutate()`, `summarise()`, `group_by()`, `ungroup()`, `case_when()`, `na_if()`
    * tidyr:: `pivot_longer()`, `pivot_wider()`
    * tibble:: `tibble()`
* other (you need to load each package to use these)
    * janitor:: `round_half_up()`


## Walkthrough video {#sec-walkthrough-wrangle .unnumbered}

There is a walkthrough video of this chapter available via [Echo360](https://echo360.org.uk/media/dc1e2869-a6c2-45d8-ab40-cb85cdb67f43/public). Please note that there may have been minor edits to the book since the video was recorded. Where there are differences, the book should always take precedence.

## Set-up {#sec-setup-wrangle}

First, create a new project for the work we'll do in this chapter named `r path("09-wrangle")`. Second, open and save and new R Markdown document named `wrangle.Rmd`, delete the welcome text and load the required packages for this chapter.

```{r setup-wrangle, message=FALSE, filename="Chapter packages"}
library(tidyverse)   # data wrangling functions
library(janitor)
```

::: {.callout-warning}
At this point in ADS it's very likely you will have the most recent version of the <pkg>tidyverse</pkg> packages but just in case, be aware that the function `case_when()` that we will use in this chapter was updated in to introduce new arguments and ways of handling `NA`. The code we provide in this chapter will only work if you have v1.1.0 or above of <pkg>dplyr</pkg> installed. To check run `packageVersion("dplyr")` and if it's below 1.1.0, install it again to update.
:::

You'll need to make a folder called "data" and download a data file into it: 
<a href="https://psyteachr.github.io/ads-v2/data/budget.csv" download>budget.csv</a>.

Download the [Data transformation cheat sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-transformation.pdf).

## Wrangling functions

Data wrangling refers to the process of cleaning, transforming, and restructuring your data to get it into the format you need for analysis and it's something you will spend an awful lot of time doing. Most `r glossary("data wrangling")` involves the reshaping functions you learned in @sec-tidy and six functions from the <pkg>dplyr</pkg> package that is loaded as part of the <pkg>tidyverse</pkg>: `select`, `filter`, `arrange`, `mutate`, `summarise`, and `group_by`. You'll remember the last two from @sec-summary, so we'll only cover them briefly.

It's worth highlighting that in this chapter we're going to cover these common functions and common uses of said functions. However, <pkg>dplyr</pkg> (and packages beyond it) has a huge number of additional wrangling functions and each function has many different arguments. Essentially, if you think you should be able to wrangle your data in a particular way that we haven't explicitly shown you, you almost certainly can, it might just take a bit of Googling to find out how. 

```{r, echo = FALSE, eval = FALSE}
set.seed(8675309)
budget <- faux::sim_design(
  between = list(
    region = c("North", "South", "East", "West"),
    product = c("widgets", "gadgets")
  ),
  within = list(
    type = c("sales", "expenses"),
    year = 2019:2020
  ),
  n = 1,
  mu = rep(c(100, 50), each = 4, times = 4),
  sd = 1000,
  plot = FALSE
) %>%
  mutate_if(is.numeric, round) %>%
  select(-id) %>%
  mutate(satisfaction_2019 = c("high", "very high", "high", "low", "very low", "neutral", "neutral", "low"),
         satisfaction_2020 = c("very high", "very high", "neutral", "low", "very low", "high", "neutral", "low"))

write_csv(budget, "data/budget.csv")
```

We'll use a small example table with the sales, expenses, and satisfaction for two years from four regions over two products. After you load the data, use `glimpse(budget)` or `View(budget)` to get familiar with the data.

```{r}
budget <- read_csv("data/budget.csv", show_col_types = FALSE)
```

```{r, echo = FALSE}
budget
```


### Select

You can select a subset of the columns (variables) in a table to make it easier to view or to prepare a table for display. You can also select columns in a new order.

#### By name or index

You can select columns by name or number (which is sometimes referred to as the column index). Selecting by number can be useful when the column names are long or complicated.

```{r}
# select single column by name
product_dat <- budget %>% select(product) 

# select single column by number
product_dat <- budget %>% select(2) 
```

You can select each column individually, separated by commas (e.g., `region, sales_2019`) but you can also select all columns from one to another by separating them with a colon (e.g., `sales_2019:expenses_2020`).

The colon notation can be much faster because you don't need to type out each individual variable name, but make sure that you know what order your columns are in and always check the output to make sure you have selected what you intended.

```{r select-name}
# select columns individually
sales2019 <- budget %>% select(region, product, sales_2019)

# select columns with colon
sales2019 <- budget %>% select(region:sales_2019)
```

You can rename columns at the same time as selecting them by setting `new_name = old_col`. 

```{r select-number}
regions <- budget %>% select(`Sales Region` = 1, 3:6)

head(regions, 2)
```

#### Un-selecting columns

You can select columns either by telling R which ones you want to keep as in the previous examples, or by specifying which ones you want to exclude by using a minus symbol to un-select columns. You can also use the colon notation to de-select columns, but to do so you need to put parentheses around the span first, e.g., `-(sales_2019:expenses_2020)`, not `-sales_2019:expenses_2020`.

```{r select-unselect}
# de-select individual columns
sales <- budget %>% select(-expenses_2019, -expenses_2020)

# de-select a range of columns
sales <- budget %>% select(-(expenses_2019:expenses_2020))
```

#### Select helpers

Finally, you can select columns based on criteria about the column names.

| function | definition |
|----------|------------|
| `starts_with()` | select columns that start with a character string|
| `ends_with()` | select columns that end with a character string |
| `contains()` | select columns that contain a character string |
| `num_range()` | select columns with a name that matches the pattern `prefix` |

```{r select-helpers, echo = FALSE}
x1 <- budget %>% select(starts_with("sales"))
x2 <- budget %>% select(ends_with("2020"))
x3 <- budget %>% select(contains("_"))
x4 <- budget %>% select(num_range("expenses_", 2019:2020))

a <- list(names(x1), names(x2), names(x3), names(x4)) %>%
  sapply(paste, collapse = ", ")

a1 <- setNames(a, c("answer", "x", "x", "x")) %>% longmcq()
a2 <- setNames(a, c("x", "answer", "x", "x")) %>% longmcq()
a3 <- setNames(a, c("x", "x", "answer", "x")) %>% longmcq()
a4 <- setNames(a, c("x", "x", "x", "answer")) %>% longmcq()
```


::: {.callout-note .try}
What are the resulting columns for these four examples?

* `budget %>% select(contains("_"))`
    `r a3`
* `budget %>% select(num_range("expenses_", 2019:2020))`
    `r a4`
* `budget %>% select(starts_with("sales"))`
    `r a1` 
* `budget %>% select(ends_with("2020"))`
    `r a2`
:::


### Filter

Whilst `select()` chooses the columns you want to retain, `filter()` chooses the rows to retain by matching row or column criteria.

You can filter by a single criterion. This criterion can be rows where a certain column's value matches a character value (e.g., "North") or a number (e.g., 9003). It can also be the result of a logical equation (e.g., keep all rows with a specific column value larger than a certain value). The criterion is checked for each row, and if the result is FALSE, the row is removed. You can reverse equations by specifying `!=` where `!` means "not".

```{r filter, eval = FALSE}
# select all rows where region equals North
budget %>% filter(region == "North")

# select all rows where expenses_2020 were exactly equal to 200
budget %>% filter(expenses_2020 == 200)

# select all rows where sales_2019 was more than 100
budget %>% filter(sales_2019 > 100)

# everything but the North
budget %>% filter(region != "North")
```

::: {.callout-warning}
Remember to use `==` and not `=` to check if two things are equivalent. A single `=` assigns the right-hand value to the left-hand variable (much like the `<-` operator).
:::


```{r filter-mcq, echo = FALSE}
class <- tibble(id = 1:4, 
               grade = c("A", "A", "C", "B"),
               score = c(95, 91, 76, 84)
               )
x1 <- which(class$grade == "A")
x2 <- which(class$score == 91)
x3 <- which(class$score < 80)
x4 <- which(class$grade != "A")

a <- list(x1, x2, x3, x4) %>%
  sapply(paste, collapse = ", ")

a1 <- setNames(a, c("answer", "x", "x", "x")) %>% mcq()
a2 <- setNames(a, c("x", "answer", "x", "x")) %>% mcq()
a3 <- setNames(a, c("x", "x", "answer", "x")) %>% mcq()
a4 <- setNames(a, c("x", "x", "x", "answer")) %>% mcq()
```


::: {.callout-note .try}
Which IDs are kept from the table below?

```{r, echo = FALSE}
class
```

* `demo %>% filter(score < 80)`
    `r a3`
* `demo %>% filter(grade == "A")`
    `r a1`
* `demo %>% filter(grade != "A")`
    `r a4`
* `demo %>% filter(score == 91)`
    `r a2`
    

:::

You can also select on multiple criteria by separating them by commas (rows will be kept if they match *all* criteria). Additionally, you can use `&` ("and") and `|` ("or") to create complex criteria.

```{r filter-symbols}
# regions and products with profit in both 2019 and 2020
profit_both <- budget %>% 
  filter(
    sales_2019 > expenses_2019,
    sales_2020 > expenses_2020
  )

# the same as above, using & instead of a comma
profit_both <- budget %>% 
  filter(
    sales_2019 > expenses_2019 &
    sales_2020 > expenses_2020
  )

# regions and products with profit in 2019 or 2020
profit_either <- budget %>% 
  filter(
    sales_2019 > expenses_2019 |
    sales_2020 > expenses_2020
  )

# 2020 profit greater than 1000
profit_1000 <- budget %>%
  filter(sales_2020 - expenses_2020 > 1000)

```

If you want the filter to retain multiple specific values in the same variable, the "match operator" (`%in%`) should be used rather than `|` (or). The `!` can also be used in combination here, but it is placed before the variable name.

```{r match-operator, eval = FALSE}
# retain any rows where region is north or south, and where product equals widget
budget %>%
  filter(region %in% c("North", "South"),
         product == "widgets")

# retain any rows where the region is not east or west, and where the product does not equal gadgets
budget %>%
  filter(!region %in% c("East", "West"),
         product != "gadgets")
```

`r glossary("Operator")`	|Name   |is TRUE if and only if
-----------|----------------------|---------------------------------
`A < B`    |less than 	          |A is less than B
`A <= B`   |less than or equal    |A is less than or equal to B
`A > B`    |greater than 	        |A is greater than B
`A >= B`   |greater than or equal |A is greater than or equal to B
`A == B`   |equivalence 	        |A exactly equals B
`A != B`   |not equal 	          |A does not exactly equal B
`A %in% B` |in 	                  |A is an element of vector B

Finally, you can also pass many other functions to filter. For example, the package <pkg>stringr</pkg> that is loaded as part of the <pkg>tidyverse</pkg> contains many different functions for working with `r glossary("string", "strings")` (character data). For example, you you use `str_detect()` to only retain rows where the customer satisfaction rating includes the word "high"

```{r eval = TRUE}
budget %>%
  filter(str_detect(satisfaction_2019, "high"))
```

Note that `str_detect()` is case sensitive so it would not return values of "High" or "HIGH". You can use the function `tolower()` or `toupper()` to convert a string to lowercase or uppercase before you search for substring if you need case-insensitive matching.

::: {.callout-warning}
`filter()` is incredibly powerful and can allow you to select very specific subsets of data. But, it is also quite dangerous because when you start combining multiple criteria and operators, it's very easy to accidentally specify something slightly different than what you intended. **Always check your output**. If you have a small dataset, then you can eyeball it to see if it looks right. With a larger dataset, you may wish to compute summary statistics or count the number of groups/observations in each variable to verify your filter is correct. There is no level of expertise in coding that can substitute knowing and checking your data. 
:::

### Arrange {#sec-arrange}

You can sort your dataset using `arrange()`. You will find yourself needing to sort data in R much less than you do in Excel, since you don't need to have rows next to each other in order to, for example, calculate group means. But `arrange()` can be useful when preparing data for display in tables. `arrange()` works on character data where it will sort alphabetically, as well as numeric data where the default is ascending order (smallest to largest). Reverse the order using `desc()`.

```{r arrange}
# arranging the table 
# first by product in alphabetical order
# then by "region" in reverse alphabetical order
budget %>%
  arrange(product, desc(region))
```


::: {.callout-note .try}
If you want to sort character data/categories in a specific order, turn the column into a `r glossary("factor")` and set the `levels` in the desired order.

```{r}
budget %>%
  mutate(region = factor(region, levels = c("North", "South", "East", "West"))) %>%
  filter(product == "gadgets") %>%
  arrange(region)
```

:::

### Mutate

The function `mutate()` allows you to add new columns or change existing ones by overwriting them by using the syntax `new_column = operation`.  You can add more than one column in the same mutate function by separating the columns with a comma. Once you make a new column, you can use it in further column definitions. For example, the creation of `profit` below uses the column `expenses`, which is created above it.

```{r mutate}
budget2 <- budget %>%
  mutate(
    sales = sales_2019 + sales_2020,
    expenses = expenses_2019 + expenses_2020,
    profit = sales - expenses,
    region = paste(region, "Office")
  )
```

`mutate()` can also be used in conjunction with other functions and Boolean operators. For example, we can add another column to `budget2` that states whether a profit was returned that year or overwrite our `product` variable as a factor. Just like when we used `r glossary("boolean expression", "Boolean expressions")` with filter, it will evaluate the equation and return TRUE or FALSE depending on whether the observation meets the criteria.

```{r}
budget2 <- budget2 %>%
  mutate(profit_category = profit > 0,
         product = as.factor(product))
```

::: {.callout-warning}
You can overwrite a column by giving a new column the same name as the old column (see `region` or `product`) above. Make sure that you mean to do this and that you aren't trying to use the old column value after you redefine it.
:::

You can also use `case_when()` to specify what values to return, rather than defaulting to TRUE or FALSE:

```{r}
budget3 <- budget2 %>%
  mutate(profit_category = case_when(profit > 0 ~ "PROFIT",
                                     profit < 0 ~ "NO PROFIT"))
```

Use it to recode values:

```{r}
# create a column where people get a bonus if customer satisfaction was overall high or very high

bonus <- budget3 %>%
  mutate(bonus_2019 = case_when(satisfaction_2019 %in% c("very low", "low", "neutral") ~ "no bonus",
                                satisfaction_2019 %in% c("high", "very high") ~ "bonus"))
```

And combine different criteria:

```{r}

# new management takes over - people only get a bonus if customer satisfaction was overall high or very high AND if a profit was returned

bonus2 <- budget3 %>%
  mutate(bonus_2020 = case_when(satisfaction_2020 == "high" & 
                                  profit_category == "PROFIT" ~ "bonus",
                                satisfaction_2020 == "very high" & 
                                  profit_category == "PROFIT" ~ "bonus",
                                .default = "No bonus")) # set all other values to "no bonus"
```

::: {.callout-tip}
Be mindful that `.default` uses `=` whilst the others use `~`. Emily has lost quite a lot of her time and sanity to not realising this. 
:::

Just like `filter()`, `mutate()` is incredibly powerful and the scope of what you can create is far beyond what we can cover in this book. 

### Summarise {#sec-dplyr-summarise}

You were introduced to the `summarise()` function in @sec-summary-summarise. This applies summary functions to an entire table (or groups, as you'll see in the next section).

Let's say we want to determine the mean sales and expenses, plus the minimum and maximum profit, for any region, product and year. First, we need to reshape the data like we learned in @sec-tidy, so that there is a column for year and one column each for sales and expenses, instead of separate columns for each year. We'll also drop the satisfaction data as we don't need it for this analysis. 

```{r}
budget4 <- budget %>%
  select(-satisfaction_2019, -satisfaction_2020) %>%
  pivot_longer(cols = sales_2019:expenses_2020,
               names_to = c("type", "year"),
               names_sep = "_",
               values_to = "value") %>%
  pivot_wider(names_from = type,
              values_from = value)

head(budget4) # check the format
```

Now we can create summary statistics for the table.

```{r}
budget4 %>%
  summarise(
    mean_sales = mean(sales),
    mean_expenses = mean(expenses),
    min_profit = min(expenses - sales),
    max_profit = max(expenses - sales)
  )
```

### Group By {#sec-dplyr-groupby}

You were introduced to the `group_by()` function in @sec-grouping. For example, you can break down the summary statistics above by year and product.

```{r}
year_prod <- budget4 %>%
  group_by(year, product) %>%
  summarise(
    mean_sales = mean(sales),
    mean_expenses = mean(expenses),
    min_profit = min(expenses - sales),
    max_profit = max(expenses - sales)
  ) %>%
  ungroup()

year_prod
```


Note that you can use the other wrangling functions on the summary table, for example: 

```{r eval = FALSE}
# arrange by maximum profit
year_prod %>%
  arrange(desc(max_profit))

# filter out gadgets
year_prod %>%
  filter(product != "gadgets")

```


::: {.callout-note .try}
How would you find out the maximum sales for each region?

```{r, results='asis', echo = FALSE}
c(answer = "<pre>budget3 %>%\n  group_by(region) %>%\n  summarise(max_sales = max(sales)</pre>",
  x = "<pre>budget3 %>%\n  group_by(region) %>%\n  summarise(max_sales = max(region)</pre>",
  x = "<pre>budget3 %>%\n  group_by(sales) %>%\n  summarise(max_sales = max(sales)</pre>",
  x = "<pre>budget3 %>%\n  group_by(sales) %>%\n  summarise(max_sales = max(region)</pre>") %>%
  sample() %>% webexercises::longmcq() %>% cat()
```

:::

You can also use `group_by()` in combination with other functions. For example, `slice_max()` returns the top N rows, ordered by a specific variable.

```{r}
# return top 3 sales
budget4 %>%
  slice_max(n = 3, order_by = sales)
```

But this can be combined with `group_by()` to return the top sales for each region:

```{r}
# return top sale for each region
budget4 %>%
  group_by(region) %>%
  slice_max(n = 1, order_by = sales)
```

## Complications

### Rounding

Let's say we want to round all the values to the nearest pound. The pattern below uses the `across()` function to apply the `round()` function to the columns from `mean_sales` to `max_profit`.

```{r}
year_prod %>%
  mutate(across(.cols = mean_sales:max_profit, 
                .fns = round))
```

If you compare this table to the one in @sec-dplyr-groupby, you'll see that the 2019 gadgets mean sales rounded up from `r year_prod$mean_sales[1]` to `r round(year_prod$mean_sales[1])`, while the mean expenses rounded down from `r year_prod$mean_expenses[1]` to `r round(year_prod$mean_expenses[1])`. What's going on!?

This may seem like a mistake, but R rounds .5 to the nearest even number, rather than always up, like you were probably taught in school. This prevents overestimation biases, since x.5 is *exactly* halfway between x and x+1, so there is no reason it should always round up.

```{r}
round(0.5)
round(1.5)
```

However, this might throw a monkey wrench into your own systems. For example, our school policy is to round up for course marks at x.5. The easiest solution is to use the `round_half_up()` function from the package <pkg>janitor</pkg>.

when you run this code, a new section will appear in the environment pane labelled "Functions". In addition to using functions from packages, you can also make your own. It's not something we are going to go into detail on in this course, but it's useful to know the functionality exists.

This should work as you'd expect.

```{r}
round_half_up(0.5)
round_half_up(1.5)
```

### Missing values {#sec-missing-values}

If you have control over your data, it is always best to keep missing values as empty cells rather than denoting missingness with a word or implausible number. If you used "missing" rather than leaving the cell empty, the entire variable would be read as character data, which means you wouldn't be able to perform mathematical operations like calculating the mean. If you use an implausible number (0 or 999 are common), then you risk these values being included in any calculations as real numbers.

However, we often don't have control over how the data come to us, so let's run through how to fix this.

#### Bad missing values

What if the South region hadn't returned their expenses (entered as 0) and the North region hadn't returned their sales data for 2020 yet, so someone entered it as "missing"?  

First, we're going to recode the data to add in the missing values

For the South data, we can use `case_when()` to set the value of expenses to 0 if the year is 2020 and region is "South", otherwise use the value from the expenses column (i.e., don't change).

```{r}
missing_bad <- budget4 %>%
  mutate(expenses = case_when(
    # set to 0 when year is 2020 and region is North
    year == 2020 & region == "South" ~ 0, 
    # otherwise, set to the value in the expenses column
    .default = expenses   
  ))
```

`r hide("Using case_when() for multiple criteria")`

The `case_when()` function allows allows you to set multiple criteria, although we're only using one non-default criterion here. It can be very useful, but takes a little practice.

The example below creates a label for each row. Notice how the label for the first row is "x < 2", even though this row also fits the second criterion "y < 4". This is because `case_when()` applies the first match to each row, even if other criteria in the function also match that row.

```{r}
data <- tibble(
  x = 1:5,
  y = 1:5
)

data %>%
  mutate(label = case_when(
    x < 2           ~ "x < 2",
    y < 4           ~ "y < 4",
    x == 5 & y == 5 ~ "both 5",
    .default        = "default"
  ))
```

`r unhide()`

For the North, we will recode these values as "missing". Since this is character data, and sales are currently numeric data, we first need to change it to a character variable.


```{r}
# set sales values to "missing" for North 2020 rows
missing_bad <- missing_bad %>%
  mutate(sales = as.character(sales),  # Convert all sales to character
         sales = case_when(year == 2020 & region == "North" ~ "missing", 
                          TRUE ~ sales))  # Set specific condition to "missing"


str(missing_bad)

```

Now, if you try to compute the mean sales, you will get an error message and the result will be `NA`.

```{r eval = TRUE, error=TRUE}
# try to compute mean sales
missing_bad %>%
  summarise(mean_sales = mean(sales))
```

#### Convert missing values to NA

To set the missing values to NA, we can use the handy function `na_if()`. We'll also need to transform `sales` back to numeric.

```{r}
missing_data <- missing_bad %>%
  mutate(
    # if expenses = 0, set as NA
    expenses = na_if(expenses,0),
    # if sales = "missing" set as NA
    sales = na_if(sales, "missing"),
    # convert to numeric
    sales = as.numeric(sales)
  )
```

Now, if we try to calculate the mean sales and profits, we get missing values for any summary value that used one of the North 2020 sales values or the South 2020 expenses.

```{r error=TRUE}
missing_data %>%
  group_by(region) %>%
  summarise(
    mean_sales = mean(sales),
    mean_expenses = mean(expenses),
    min_profit = min(expenses - sales),
    max_profit = max(expenses - sales),
    .groups = "drop")
```

#### Ignore missing values

This is because `NA` basically means "I don't know", and the sum of 100 and "I don't know" is "I don't know", not 100. However, when you're calculating means, you often want to just ignore missing values. Set `na.rm = TRUE` in the summary function to remove missing values before calculating.

```{r}
missing_data %>%
  group_by(region) %>%
  summarise(
    mean_sales = mean(sales, na.rm = TRUE),
    mean_expenses = mean(expenses, na.rm = TRUE),
    min_profit = min(expenses - sales, na.rm = TRUE),
    max_profit = max(expenses - sales, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Count missing values

If you want to find out how many missing or non-missing values there are in a column, use the `is.na()` function to get a `r glossary("logical")` vector of whether or not each value is missing, and use `sum()` to count how many values are TRUE or `mean()` to calculate the proportion of TRUE values.

```{r}
missing_data %>%
  group_by(year, product) %>%
  summarise(
    n_valid = sum(!is.na(sales)),
    n_missing = sum(is.na(sales)),
    prop_missing = mean(is.na(sales)),
    .groups = "drop"
  )
```

#### Omit missing values

You may also want to remove rows that have missing values and only work from complete datasets. `drop_na()` will remove any row that has a missing observation. You can use `drop_na()` on the entire dataset which will remove any row that has *any* missing value, or you can specify to only remove rows that are missing a specific value.

```{r}
# remove any rows with any missing values
complete_data <- missing_data %>%
  drop_na()

# remove any rows that are missing a value for sales
complete_sales <- missing_data %>%
  drop_na(sales)

```

Missing data can be quite difficult to deal with depending on how it is represented. As always, no amount of coding expertise can make up for not understanding the structure and idiosyncrasies of your data. 

## Exercises {#sec-together-wrangle}

Let's try some exercises using a dataset you already encountered in @sec-viz so that you can see how much more you're able to do with the data now.

* Save your current Markdown, close it, and open a new Rmd named "survey_data_mad_skillz".
* In the set-up code chunk, load the tidyverse, then load the dataset from https://psyteachr.github.io/ads-v2/data/survey_data.csv into an object named `survey_data`.
* Use your method of choice to review the dataset and familiarise yourself with its structure.

```{r webex.hide = TRUE}
# from https://www.kaggle.com/kyanyoga/sample-sales-data
library(tidyverse)
survey_data <- read_csv("https://psyteachr.github.io/ads-v2/data/survey_data.csv")

```

### Creating new categories

Employees 1-5 were trained by Michael and employees 6-10 were trained by Dwight. 

* Create a new column named `trainer` that lists the trainer for each employee.
* Then, calculate the average satisfaction scores for employees trained by each trainer and visualise the satisfaction scores for each in whatever way you think best.

` r hide("Hint")`
To add the trainer column you can use `case_when()` and specify multiple criteria (e.g., if the employee is 1-5, Michael, if the employee is 6-10 Dwight)
` r unhide()`

```{r webex.hide = TRUE}
# case_when() method
survey_data <- survey_data %>%
  mutate(trainer = case_when(employee_id %in% c("E01", "E02", "E03", "E04", "E05") ~ "Michael",
                             employee_id %in% c("E06", "E07", "E08", "E09", "E10") ~ "Dwight"))


# mean satisfaction scores
survey_data %>%
  group_by(trainer) %>%
  summarise(mean_satisfaction = mean(satisfaction))

# possible visualisation 

ggplot(survey_data, aes(x = satisfaction, fill = trainer)) +
  geom_histogram(binwidth = 1, show.legend = FALSE, colour = "black") +
  facet_wrap(~trainer) +
  labs(title = "Satisfaction scores by employee trainer")
```

### Filter by calculated score

First, calculate the average wait time and store this in an object named `mean_wait`. This should be a single value rather than a table.

`r hide("Hint")`
There are multiple ways to achieve this. You could create the table and then pull out the single value, or just calculate the single value.
`r unhide()`

```{r webex.hide = TRUE}

# method 1 - tidyverse
mean_wait <- survey_data %>%
  summarise(mean_wait = mean(wait_time)) %>%
  pull(mean_wait)

# method 2 - base R
mean_wait <- mean(survey_data$wait_time)

```

Now create a dataset named `long_wait` that just contains data from customers who waited more than the average wait time.

```{r webex.hide = TRUE}
long_wait <- survey_data %>%
  filter(wait_time > mean_wait)
```

Create a visualisation that shows how many customers waited more than the average wait time for each employee.

```{r webex.hide = TRUE}
long_wait %>%
  ggplot(aes(x = employee_id)) +
  geom_bar()
```

### Multiple critera

Now, add a column to `survey_data` named `follow_up` that flags whether a customer should be followed up with a courtesy phone call. Your company is short-staffed so only customers that meet all three of the following criteria should be followed-up:

* Their wait time should be above the average **for all calls**
* Their call time should be above the average **for their category**
* Their satisfaction should be less than three 3. 

This is quite complicated and there are multiple ways to achieve the desired outcome. Some approaches may need other functions that were covered in previous chapters and you may need to create intermediate objects.

Call the final object `follow_data` and keep only the customer ID, employee ID, trainer, and follow up columns. 

```{r webex.hide = TRUE}

# this is one possible solution, there are many other valid approaches 

# calculate mean wait time across all calls
mean_wait <- mean(survey_data$wait_time)

# calculate mean call time for each category
follow_data <- survey_data %>%
  group_by(issue_category) %>%
  summarise(mean_call = mean(call_time)) %>%
#then join it to the survey data  
  left_join(survey_data, by = "issue_category") %>%
# then add on the column
  mutate(follow_up = case_when(wait_time > mean_wait & 
                               call_time > mean_call & 
                               satisfaction < 3 ~ "yes",
                               .default = "no")) %>%
  select(caller_id, employee_id, trainer, follow_up)
```

For all of the above, write code that stores the answer as a single value, so that you could easily use it in inline coding.

How many customers need to be followed up:

* In total? `r fitb("120")`
* From calls by employee 06? `r fitb("16")`
* From calls by employees trained by Michael `r fitb("65")`
* From calls by employees trained by Dwight `r fitb("55")`

`r hide("Hint")`
`group_by %>% count() %>% filter() %>% pull()
`r unhide()`

Which employee needs to make the largest number of follow-up courtesy calls? `r fitb("E02")`

`r hide("Hint")`
As above but add in an `ungroup()` and `slice_max()` along the way. 
`r unhide()`

```{r webex.hide = TRUE}
# in total
follow_data %>%
  group_by(follow_up) %>%
  count()%>%
  filter(follow_up == "yes") %>%
  pull(n)

# by employee 6
follow_data %>%
  group_by(follow_up, employee_id) %>%
  count() %>%
  filter(employee_id == "E06",
         follow_up == "yes") %>%
  pull(n)

# by michael
follow_data %>%
  group_by(follow_up, trainer) %>%
  count() %>%
  filter(trainer == "Michael",
         follow_up == "yes") %>%
  pull(n)

# by dwight
follow_data %>%
  group_by(follow_up, trainer) %>%
  count() %>%
  filter(trainer == "Dwight",
         follow_up == "yes") %>%
  pull(n)

# most follow-ups needed
follow_data %>%
  group_by(follow_up, employee_id) %>%
  count() %>%
  ungroup() %>%
  filter(follow_up == "yes") %>%
  slice_max(n = 1, order_by = n) %>%
  pull(employee_id)

```

### Original insight

In preparation for the final summative assessment, explore the data to provide one original insight of your own.

### Share on Teams

Once you're done, share your knitted html file and the Rmd file for the exercises on Teams in the Week 09 channel.


## Glossary {#sec-glossary-wrangle}

```{r, echo = FALSE, results='asis'}
glossary_table(as_kable = FALSE) |> 
  kableExtra::kable(row.names = FALSE, escape = FALSE) |>
  unclass() |> cat()
```

## Further resources {#sec-resources-wrangle}

* [Data transformation cheat sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-transformation.pdf)
* [Chapter 18: Missing Data](https://r4ds.hadley.nz/missing-values) in *R for Data Science*
* [Chapter 3: Data Transformation ](https://r4ds.hadley.nz/data-transform) in *R for Data Science*
* [Chapter 25: Functions](https://r4ds.hadley.nz/functions) in *R for Data Science*
* [Introduction to stringr](https://stringr.tidyverse.org/articles/stringr.html)




