# Analysis Journey 2: Simple Linear Regression {#journey-02-wrangling}

Welcome to the second data analysis journey. We have designed these chapters as a bridge between the structured learning in the core chapters and your assessments. We present you with a new data set, show you what the end product should look like, and see if you can apply your data wrangling, visualisation, and/or analysis skills to get there. 

As you gain independence, this is the crucial skill. Data analysis is all about seeing the data you have available to you and identifying what the end product needs to be to apply your visualisation and analysis techniques. You can then mentally (or physically) create a checklist of tasks to work backwards to get there. There might be a lot of trial and error as you try one thing, it does not quite work, so you go back and try something else. If you get stuck though, we have a range of hints and task lists you can unhide, then the solution to check your attempts against. 

In this second data analysis journey, we focus on inferential statistics after some data wrangling to apply all the skills you developed from Chapter 1 to Chapter 9. 

## Task preparation

### Introduction to the data set 

For this task, we are using open data from @binfet_importance_2022, where the authors used the data set to write a separate article on repurposing it for statistics education [@evans_repurposing_2023], inspiring us to use it in the chapter here. The abstract of their article is: 

> Researchers have claimed that canine-assisted interventions (CAIs) contribute significantly to bolstering participants' wellbeing, yet the mechanisms within interactions have received little empirical attention. The aim of this study was to assess the impact of client–canine contact on wellbeing outcomes in a sample of 284 undergraduate college students (77% female; 21% male, 2% non-binary). Participants self-selected to participate and were randomly assigned to one of two canine interaction treatment conditions (touch or no touch) or to a handler-only condition with no therapy dog present. To assess self-reports of wellbeing, measures of flourishing, positive and negative affect, social connectedness, happiness, integration into the campus community, stress, homesickness, and loneliness were administered. Exploratory analyses were conducted to assess whether these wellbeing measures could be considered as measuring a unidimensional construct. This included both reliability analysis and exploratory factor analysis. Based on the results of these analyses we created a composite measure using participant scores on a latent factor. We then conducted the tests of the four hypotheses using these factor scores. Results indicate that participants across all conditions experienced enhanced wellbeing on several measures; however, only those in the direct contact condition reported significant improvements on all measures of wellbeing. Additionally, direct interactions with therapy dogs through touch elicited greater wellbeing benefits than did no touch/indirect interactions or interactions with only a dog handler. Similarly, analyses using scores on the wellbeing factor indicated significant improvement in wellbeing across all conditions (handler-only, *d*=0.18, *p*=0.041; indirect, *d*=0.38, *p*<0.001; direct, *d*=0.78, *p*<0.001), with more benefit when a dog was present (*d*=0.20, *p*<0.001), and the most benefit coming from physical contact with the dog (*d*=0.13, *p*=0.002). The findings hold implications for post-secondary wellbeing programs as well as the organization and delivery of CAIs.

In summary, they were interested in the effect of therapy dogs on well-being in undergraduate students. Participants were randomly allocated to one of three groups: 

1. Canine interaction touching the dogs (Direct). 

2. Canine interaction not touching the dogs (Indirect). 

3. Handler-only with no dogs present (Control). 

They measured 9 outcomes before and after the intervention including social connectedness, stress, and loneliness. For this journey chapter, we will focus on a constrained set of variables and analyses so it does not take forever, but the process would apply to all the outcomes. The authors posed three hypotheses which we will test after some data wrangling:

1. All treatment groups would have significantly higher measures of well-being and lower measures of ill-being after treatment. 

2. The treatment groups that interact with dogs would have significantly higher measures of well-being and lower measures of ill-being compared to the handler-only treatment. 

3. Direct contact with a therapy dog would yield greater benefits than indirect contact treatment.

### Organising your files and project for the task

Before we can get started, you need to organise your files and project for the task, so your working directory is in order.

1. In your folder for research methods and the book `ResearchMethods1_2/Quant_Fundamentals`, create a new folder for the data analysis journey called `Journey_02_regression`. Within `Journey_02_regression`, create two new folders called `data` and `figures`.

2. Create an R Project for `Journey_02_regression` as an existing directory for your chapter folder. This should now be your working directory.

3. Create a new R Markdown document and give it a sensible title describing the chapter, such as `Analysis Journey 2 - Simple Linear Regression`. Delete everything below line 10 so you have a blank file to work with and save the file in your `Journey_02_regression` folder. 

4. We are working with a new data set, so please save the following data file: [Evans_2023_raw.csv](data/Evans_2023_raw.csv). Right click the link and select "save link as", or clicking the link will save the files to your Downloads. Make sure that you save the file as ".csv". Save or copy the file to your `data/` folder within `Journey_02_regression`.

You are now ready to start working on the task! 

## Overview

### Load <pkg>tidyverse</pkg> and read the data file

Before we explore what wrangling we need to do, complete the following task list and check the solution if you are stuck. 

::: {.callout-tip}
#### Try this
Complete the following steps:

1. Load the <pkg>tidyverse</pkg> package. 
    
2. Read the data file `data/Evans_2023_raw.csv` to the object name `evans_data`.
:::

::: {.callout-tip collapse="true"}
#### Show me the solution
You should have the following in a code chunk: 

```{r message=FALSE, warning=FALSE}
# load the relevant packages
library(tidyverse)

# Read the Evans_2023_raw.csv file 
evans_data <- read_csv("data/Evans_2023_raw.csv")
```
:::

### Explore `evans_data`

In `evans_data`, we have the participant ID (`RID`), several demographic variables, and pre- and post-test items for stress, loneliness, and social connectedness. There are 88 variables which would take up loads of space, so we are just showing a preview of the first 20 here. If you use `glimpse()`, you will see all 88. 

```{r echo=FALSE}
evans_data %>% 
  select(1:20) %>% 
  glimpse()
```

The columns (variables) we have in the data set are:

| Variable       |       Type                       |           Description          |
|:--------------:|:---------------------------------|:-------------------------------|
| RID | `r typeof(evans_data$RID)`| Participant ID number. |
| GroupAssignment | `r typeof(evans_data$GroupAssignment)`| Randomly allocated study group: Control, Indirect, Direct. |
| Age_Yrs | `r typeof(evans_data$Age_Yrs)`| Age in years. |
| Year_of_Study | `r typeof(evans_data$Year_of_Study)`| Participant's year in college: First (1), Second (2), Third (3), Fourth (4), Fifth or more (5). |
| Live_Pets | `r typeof(evans_data$Live_Pets)`| Does the participant have a pet back at home: Pet back home (1), no pet back home (2). |
| Consumer_BARK | `r typeof(evans_data$Consumer_BARK)`| Is the participant a low (1), medium (2), or high (3) consumer of the BARK program - the therapy dog service. |
| S1_1 | `r typeof(evans_data$S1_1)`| Stress scale pre-test, 1 item, 1 (not at all stressed) to 5 (very stressed). |
| L1_1 to L1_20 | `r typeof(evans_data$L1_1)`| Loneliness scale pre-test, 20 items, 1 (never) to 4 (often). |
| SC1_1 to SC1_20 | `r typeof(evans_data$SC1_1)`| Social connectedness scale pre-test, 20 items, 1 (strongly disagree) to 6 (strongly agree). |
| S2_1 | `r typeof(evans_data$S2_1)`| Stress scale post-test, 1 item. |
| L2_1 to L2_20 | `r typeof(evans_data$L2_1)`| Loneliness scale post-test, 20 items. |
| SC2_1 to SC2_20 | `r typeof(evans_data$SC2_1)`| Social connectedness scale post-test, 20 items, 1 (strongly disagree) to 6 (strongly agree).|

::: {.callout-tip}
#### Try this
Now we have introduced the data set, explore them using different methods we introduced. For example, opening the data object as a tab to scroll around, explore with `glimpse()`, or even try plotting some of the individual variables to see what they look like.
:::

## Wrangling 

```{r echo=FALSE, warning=FALSE, message=FALSE}
evans_clean <- evans_data %>% 
  mutate(Live_Pets = case_match(Live_Pets,
                                1 ~ "Has a pet back home",
                                2 ~ "Does not have a pet back home"),
         Consumer_BARK = case_match(Consumer_BARK,
                                    1 ~ "Low",
                                    2 ~ "Medium",
                                    3 ~ "High"),
         across(.cols = c(L1_1, L1_5, L1_6, L1_9, L1_10, L1_15, L1_16, L1_19, L1_20,
                          L2_1, L2_5, L2_6, L2_9, L2_10, L2_15, L2_16, L2_19, L2_20),
                .fns = ~ 5 - .x),
         across(.cols = c(SC1_3, SC1_6, SC1_7, SC1_9, SC1_11, SC1_13, SC1_15, SC1_17, SC1_18, SC1_20, 
                          SC2_3, SC2_6, SC2_7, SC2_9, SC2_11, SC2_13, SC2_15, SC2_17, SC2_18, SC2_20),
                .fns = ~ 7 - .x))

lonely_pre <- evans_clean %>% 
  pivot_longer(cols = L1_1:L1_20, 
               names_to = "Item", 
               values_to = "Response") %>% 
  group_by(RID) %>% 
  summarise(lonely_pre = mean(Response, na.rm = TRUE))

lonely_post <- evans_clean %>% 
  pivot_longer(cols = L2_1:L2_20, 
               names_to = "Item", 
               values_to = "Response") %>% 
  group_by(RID) %>% 
  summarise(lonely_post = mean(Response, na.rm = TRUE))

social_pre <- evans_clean %>% 
  pivot_longer(cols = SC1_1:SC1_20, 
               names_to = "Item", 
               values_to = "Response") %>% 
  group_by(RID) %>% 
  summarise(social_pre = mean(Response, na.rm = TRUE))

social_post <- evans_clean %>% 
  pivot_longer(cols = SC2_1:SC2_20, 
               names_to = "Item", 
               values_to = "Response") %>% 
  group_by(RID) %>% 
  summarise(social_post = mean(Response, na.rm = TRUE))

evans_wide <- evans_clean %>% 
  inner_join(lonely_pre) %>% 
  inner_join(lonely_post) %>% 
  inner_join(social_pre) %>% 
  inner_join(social_post) %>% 
  select(RID:Consumer_BARK, 
         stress_pre = S1_1, 
         stress_post = S2_1, 
         lonely_pre:social_post)
```

We are going to show you a preview of the starting data set and the end product we are aiming for. For the raw data, we have limited this to the first 20 rows again just so it does not take up the whole page, but if you use `glimpse()` you will see all 88 variables.

::: {.panel-tabset}
#### Raw data

```{r echo=FALSE}
evans_data %>% 
  select(1:20) %>% 
  glimpse()
```

#### Wrangled data

```{r echo=FALSE}
glimpse(evans_wide)
```
:::

::: {.callout-tip}
#### Try this

Before we give you a task list, try and switch between the raw data and the wrangled data. Make a list of all the differences you can see between the two data objects. 

1. Do the values of variables change from numbers? How might you recode them using the code book above?  

2. Looking at the codebook, are some variables the same but renamed? 

3. Looking at the codebook, have we calculated the mean of all the items for a scale?

Try and wrangle the data based on all the differences you notice to create a new object `evans_wide`. 

For one hint, unless you read the original paper, there are a bunch of items that first need reverse coding you would not know about:

- Loneliness pre-test: L1_1, L1_5, L1_6, L1_9, L1_10, L1_15, L1_16, L1_19, L1_20.

- Loneliness post-test: L2_1, L2_5, L2_6, L2_9, L2_10, L2_15, L2_16, L2_19, L2_20.
                          
- Social connectedness pre-test: SC1_3, SC1_6, SC1_7, SC1_9, SC1_11, SC1_13, SC1_15, SC1_17, SC1_18, SC1_20.

- Social connectedness post-test: SC2_3, SC2_6, SC2_7, SC2_9, SC2_11, SC2_13, SC2_15, SC2_17, SC2_18, SC2_20.

When you get as far as you can, check the task list which explains all the steps we applied, but not how to do them. Then, you can check the solution for our code. 

:::

### Task list

::: {.callout-note collapse="true"} 
#### Show me the task list

These are all the steps we applied to create the wrangled data object:

1. Recode `Live_Pets` to the two labels outlined in the code book. 

2. Recode `Consumer_BARK` to the three labels outlined in the code book. 

3. Reverse code the loneliness and social connectedness items outlined above. Think of previous examples where we explained reverse coding for how you can do this efficiently. 

As one extra piece of advice if you do not want to recode 40 variables one by one, there is a more advanced function you can use within `mutate()`. The function `across()` lets you apply a function or calculation to several columns at once. For example, if we wanted to reverse score items on a 4-point scale, it would look like the following: 

```{r eval=FALSE}
mutate(across(.cols = c(column1, column2...), 
              .fns = ~ 5 - .x))
```

In `.cols`, we enter all the columns we want to apply the function to. 

In `.fns` after the `=`, we add the function we want to apply to all the columns we selected. The code is a little awkward as we have a tilde `~`, here the calculation we want to apply, and `.x` in place of the column name. You could summarise it as: for all the columns I select, subtract each value from 5. Once you get used to the format, `across()` is really helpful when you want to do the same thing to multiple columns. 

4. After reverse coding the items, calculate the subscale mean scores for loneliness and social connectedness. You must do this twice per scale, as we have the 20 items for the pre-test and 20 items for the post-test per scale. 

5. If you calculated the subscale mean scores individually, join them back to the `evans_clean` object you mutated. 

6. Select the following columns: 

    - `RID` to `Consumer_BARK`. 

    - Rename `S1_1` to `stress_pre`.
    
    - Rename `S2_1` to `stress_post`.
    
    - Select your four subscale mean score variables. 

Remember: If it's easier for you to complete steps with longer but accurate code, there is nothing wrong with that. You recognise ways to make your code more efficient over time. 

:::

### Solution

::: {.callout-caution collapse="true"} 
#### Show me the solution

This is the code we used to create the new object `evans_wide` using the original object `evans_data`. As long as you get the same end result, the exact code is not important. In coding, there are multiple ways of getting to the same end result. Maybe you found a more efficient way to complete some of the steps compared to us. Maybe your code was a little longer. As long as it worked, that is the most important thing.

```{r eval=FALSE}
# Initial cleaning step to recode pets and BARK
# then reverse code a bunch of items 
evans_clean <- evans_data %>% 
  mutate(Live_Pets = case_match(Live_Pets,
                                1 ~ "Has a pet back home",
                                2 ~ "Does not have a pet back home"),
         Consumer_BARK = case_match(Consumer_BARK,
                                    1 ~ "Low",
                                    2 ~ "Medium",
                                    3 ~ "High"),
         # across works with mutate to apply the same function to several columns
         # So, take all the loneliness items to reverse code, then subtract them from 5
         across(.cols = c(L1_1, L1_5, L1_6, L1_9, L1_10, L1_15, L1_16, L1_19, L1_20,
                          L2_1, L2_5, L2_6, L2_9, L2_10, L2_15, L2_16, L2_19, L2_20),
                .fns = ~ 5 - .x),
         # take all the connectedness items to reverse code, then subtract them from 7
         across(.cols = c(SC1_3, SC1_6, SC1_7, SC1_9, SC1_11, SC1_13, SC1_15, SC1_17, SC1_18, SC1_20, 
                          SC2_3, SC2_6, SC2_7, SC2_9, SC2_11, SC2_13, SC2_15, SC2_17, SC2_18, SC2_20),
                .fns = ~ 7 - .x))

# There are more elegant ways around this, but for each set, 
# take the 20 items, group by participant ID, and calculate the mean, ignoring missing values
lonely_pre <- evans_clean %>% 
  pivot_longer(cols = L1_1:L1_20, 
               names_to = "Item", 
               values_to = "Response") %>% 
  group_by(RID) %>% 
  summarise(lonely_pre = mean(Response, na.rm = TRUE))

# Same thing for post scores
lonely_post <- evans_clean %>% 
  pivot_longer(cols = L2_1:L2_20, 
               names_to = "Item", 
               values_to = "Response") %>% 
  group_by(RID) %>% 
  summarise(lonely_post = mean(Response, na.rm = TRUE))

# take the 20 items, group by participant ID, and calculate the mean, ignoring missing values
social_pre <- evans_clean %>% 
  pivot_longer(cols = SC1_1:SC1_20, 
               names_to = "Item", 
               values_to = "Response") %>% 
  group_by(RID) %>% 
  summarise(social_pre = mean(Response, na.rm = TRUE))

# Same thing for post scores
social_post <- evans_clean %>% 
  pivot_longer(cols = SC2_1:SC2_20, 
               names_to = "Item", 
               values_to = "Response") %>% 
  group_by(RID) %>% 
  summarise(social_post = mean(Response, na.rm = TRUE))

# join all four summary values to main data
# select just the key variables we need
# rename the two stress items
evans_wide <- evans_clean %>% 
  inner_join(lonely_pre) %>% 
  inner_join(lonely_post) %>% 
  inner_join(social_pre) %>% 
  inner_join(social_post) %>% 
  select(RID:Consumer_BARK, 
         stress_pre = S1_1, 
         stress_post = S2_1, 
         lonely_pre:social_post)

```
:::

## Summarising/visualising 

You should now have an object called `evans_wide` containing 12 variables. If you struggled completing the wrangling steps, you can copy the code from the solution to follow along from this point. In this section, we will calculate some summary statistics and plot the data to see what we can learn. We present you with a list of questions to answer using your wrangling and visualisation skills, interspersed with the solutions to check if you are stuck. 

### Demographics

For demographics, we will recreate some values from Table 1 from @binfet_importance_2022.

1. How many participants were in each group for `GroupAssignment`? 

    - `r fitb(94)` Control
    
    - `r fitb(95)` Direct
    
    - `r fitb(95)` Indirect

::: {.callout-caution collapse="true"} 
#### Show me the solution

This nicely reproduces their values. 

```{r}
evans_wide %>% 
  count(GroupAssignment)
```
:::

2. To 2 decimals, what was the mean and standard deviation age per group? 

    - Control: *M* = `r fitb(19.95)`, *SD* = `r fitb(2.89)`. 
    
    - Direct: *M* = `r fitb(19.77)`, *SD* = `r fitb(1.94)`.
    
    - Indirect: *M* = `r fitb(19.95)`, *SD* = `r fitb(2.23)`.

::: {.callout-caution collapse="true"} 
#### Show me the solution

Weirdly, this reproduces the standard deviations from Table 1, but not the means. 

```{r}
evans_wide %>% 
  group_by(GroupAssignment) %>% 
  summarise(mean_age = round(mean(Age_Yrs), 2),
            sd_age = round(sd(Age_Yrs), 2))
```
:::

3. How many participants in each group have a pet at home? 

    - `r fitb(21)` Control
    
    - `r fitb(30)` Direct
    
    - `r fitb(34)` Indirect

::: {.callout-caution collapse="true"} 
#### Show me the solution

This nicely reproduces their values. 

```{r}
evans_wide %>% 
  count(GroupAssignment, Live_Pets)
```
:::

4. This is not part of their article, but one interesting question might be how many people have a pet at home (`Live_Pets`) against how frequently they use the BARK program (`Consumer_BARK`). Try and recreate the following bar plot to visualise this as close as possible. We have intentionally used some features we have not covered in the data visualisation chapters to get you problem solving. For one hint though, we used option D for the viridis colour scheme. 

```{r echo=FALSE}
evans_wide %>% 
  drop_na(Live_Pets, Consumer_BARK) %>% 
  mutate(Consumer_BARK = factor(Consumer_BARK, 
                                levels = c("Low", "Medium", "High"))) %>% 
  ggplot(aes(x = Live_Pets, fill = Consumer_BARK)) + 
  geom_bar(position = "dodge") + 
  labs(x = "Pets in Home", 
       y = "Frequency") + 
  scale_fill_viridis_d(option = "D", 
                       name = "BARK Program User") + 
  theme_classic()
```

::: {.callout-caution collapse="true"} 
#### Show me the solution

The new features we hoped you found independently were: 

- Setting the factor order to show `Consumer_BARK` as low, medium, then high. 

- Set `position = "dodge"` to avoid the stacked bar chart. 

- Edited the legend title by using the `name` argument in the `scale_fill` layer. 

```{r eval=FALSE}
evans_wide %>% 
  drop_na(Live_Pets, Consumer_BARK) %>% 
  mutate(Consumer_BARK = factor(Consumer_BARK, 
                                levels = c("Low", "Medium", "High"))) %>% 
  ggplot(aes(x = Live_Pets, fill = Consumer_BARK)) + 
  geom_bar(position = "dodge") + 
  labs(x = "Pets in Home", 
       y = "Frequency") + 
  scale_fill_viridis_d(option = "D", 
                       name = "BARK Program User") + 
  theme_classic()
```
:::

### Wellbeing measures

For wellbeing and ill-being measures, we will recreate some values from Table 2 from @binfet_importance_2022.

5. If you calculate the mean and standard deviation of each variable per group, answer the following questions: 

    - Which group has the lowest post-test stress value? `r mcq(sample(c("Control", "Indirect", answer = "Direct")))`
    
    - Which group has the lowest post-test loneliness value? `r mcq(sample(c("Control", "Indirect", answer = "Direct")))`
    
    - Which group has the lowest post-test social connectedness value? `r mcq(sample(c(answer = "Control", "Indirect", "Direct")))`

::: {.callout-caution collapse="true"} 
#### Show me the solution

This table calculates the mean and standard score for each variable per group. You can use this to answer the questions above. 

```{r}
evans_wide %>% 
  pivot_longer(cols = stress_pre:social_post, 
               names_to = "Variable", 
               values_to = "Value") %>% 
  group_by(GroupAssignment, Variable) %>% 
  summarise(mean_score = round(mean(Value), 2), 
            sd_score = round(sd(Value), 2))

```
:::

6. Create a scatterplot of the relationship between post-test social connectedness and loneliness. The relationship between the two variables is `r mcq(sample(c("positive", answer = "negative")))`, meaning that as social connectedness increases, we expect loneliness to `r mcq(sample(c("increase", answer = "decrease")))`.

::: {.callout-caution collapse="true"} 
#### Show me the solution

The code here creates a scatterplot for the relationship between post-test social connectedness and loneliness. 

```{r}
evans_wide %>% 
  ggplot(aes(x = lonely_post, y = social_post)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  theme_classic() + 
  labs(x = "Loneliness Post-test", y = "Social Connectedness Post-test")
```
:::

7. As we start to think about inferential statistics, we can plot the difference between pre- and post-test for each group. For this question, try and recreate the boxplot to visualise stress. Hint: think about if you need to restructure the data, and how you can present the conditions in the appropriate order. 

```{r echo=FALSE}
evans_wide %>% 
  pivot_longer(cols = stress_pre:stress_post, 
               names_to = "Stress", 
               values_to = "Value") %>% 
  mutate(Stress = factor(Stress,
                         levels = c("stress_pre", "stress_post"), 
                         labels = c("Pre-test", "Post-test"))) %>% 
  ggplot(aes(x = GroupAssignment, y = Value, fill = Stress)) + 
  geom_boxplot(alpha = 0.7) + 
  labs(x = "Group Assignment", y = "Stress Scale") + 
  scale_fill_viridis_d(option = "E", 
                       name = "Time") + 
  theme_classic()
```

::: {.callout-caution collapse="true"} 
#### Show me the solution

There are two key steps before you can plot the data. First, you need to restructure the data to long form to get one variable for pre- and post-test. Second, post comes before pre due to alphabetical order, so you need to create a factor to specify the order here. 

```{r eval=FALSE}
evans_wide %>% 
  pivot_longer(cols = stress_pre:stress_post, 
               names_to = "Stress", 
               values_to = "Value") %>% 
  mutate(Stress = factor(Stress,
                         levels = c("stress_pre", "stress_post"), 
                         labels = c("Pre-test", "Post-test"))) %>% 
  ggplot(aes(x = GroupAssignment, y = Value, fill = Stress)) + 
  geom_boxplot(alpha = 0.7) + 
  labs(x = "Group Assignment", y = "Stress Scale") + 
  scale_fill_viridis_d(option = "E", 
                       name = "Time") + 
  theme_classic()
```
:::

8. Try and recreate the violin-boxplot to visualise loneliness. Hint: remember how to align the different elements from Chapter 7. 

```{r warning=FALSE, message=FALSE, echo=FALSE}
# specify as an object, so we only change it in one place
dodge_value <- 0.9

evans_wide %>% 
  pivot_longer(cols = lonely_pre:lonely_post, 
               names_to = "Lonely", 
               values_to = "Value") %>% 
   mutate(Lonely = factor(Lonely,
                          levels = c("lonely_pre", "lonely_post"), 
                          labels = c("Pre-test", "Post-test"))) %>% 
  ggplot(aes(x = GroupAssignment, y = Value, fill = Lonely)) + 
  geom_violin(alpha = 0.5) + 
  geom_boxplot(width = 0.2, 
               alpha = 0.7,
               fatten = NULL,
               position = position_dodge(dodge_value)) + 
  stat_summary(fun = "mean", 
               geom = "point",
               position = position_dodge(dodge_value)) +
  stat_summary(fun.data = "mean_cl_boot", 
               geom = "errorbar", 
               width = .1,
               position = position_dodge(dodge_value)) +
  scale_fill_viridis_d(option = "E", 
                       name = "Time") + 
  labs(x = "Group Assignment", y = "Loneliness Scale") + 
  theme_classic()
```

::: {.callout-caution collapse="true"} 
#### Show me the solution

The same hints apply as question 7 as you need to restructure the data and create a new factor order so pre comes before post. As we have two grouping variables, we must specify a constant position dodge value so it does not plot weird. 

```{r eval=FALSE}
# specify as an object, so we only change it in one place
dodge_value <- 0.9

evans_wide %>% 
  pivot_longer(cols = lonely_pre:lonely_post, 
               names_to = "Lonely", 
               values_to = "Value") %>% 
   mutate(Lonely = factor(Lonely,
                          levels = c("lonely_pre", "lonely_post"), 
                          labels = c("Pre-test", "Post-test"))) %>% 
  ggplot(aes(x = GroupAssignment, y = Value, fill = Lonely)) + 
  geom_violin(alpha = 0.5) + 
  geom_boxplot(width = 0.2, 
               alpha = 0.7,
               fatten = NULL,
               position = position_dodge(dodge_value)) + 
  stat_summary(fun = "mean", 
               geom = "point",
               position = position_dodge(dodge_value)) +
  stat_summary(fun.data = "mean_cl_boot", 
               geom = "errorbar", 
               width = .1,
               position = position_dodge(dodge_value)) +
  scale_fill_viridis_d(option = "E", 
                       name = "Time") + 
  labs(x = "Group Assignment", y = "Loneliness Scale") + 
  theme_classic()
```
:::

## Analysing 

Finally, we turn to analysis for inferential statistics. In @evans_repurposing_2023, they organise the analyses from @binfet_importance_2022 into three hypotheses. You have not covered all the skills in Research Methods 1 to reproduce their analyses exactly, so we will work around an adapted set based on what we currently expect of you. 

We present each analysis as an overview so you can think about what techniques from Chapters 8 and 9 would address it, give you instructions on what analysis we have in mind if you need guidance, then present the solution. We focus on one outcome per hypothesis, but once you are confident you are applying and interpreting the appropriate techniques, why not try the other outcomes yourself? 

### Hypothesis 1

9. Hypothesis 1 is that each treatment group will increase measures of well-being (social connectedness) and decrease measures of ill-being (stress and loneliness). For this question, we focus on stress, so we expected stress to decrease. 

::: {.callout-tip}
#### Try this
There are two ways you could approach this. Either as the whole sample, or like @evans_repurposing_2023 present to focus on each group separately. Think about what techniques from Chapters 8 and 9 would let you test the hypothesis that each treatment group will decrease stress at post-test compared to pre-test.
:::

::: {.callout-note collapse="true"} 
#### Show me the task list

For the solution below, we are focusing on the whole sample to test whether everyone decreased stress in post-test, but you could approach it by separating the data into the three groups and then testing for the difference per group. 

To test whether an outcome decreases between conditions in the same participants, we need a model suitable for a within-subjects design. In Chapter 9, we covered in [the bonus section](#09-bonus) how to test the difference in conditions either through a paired samples-test or linear model with fixed intercept on the difference score. 

1. Calculate the difference between stress pre- and post-test. 

2. Fit a linear model on the difference score with a fixed intercept and no predictor. 

3. Look at the intercept, is it positive or negative? If you calculated pre-test minus post-test, you would be looking for a postive difference as we expect lower stress at post-test. For hypothesis testing, is the *p*-value lower than alpha? 

4. What is the effect size? How much did stress change from pre-test to post-test? What is the confidence interval around the effect size? 

:::

::: {.callout-caution collapse="true"} 
#### Show me the solution

In the code below, we first calculate a difference score by substracting stress post-test from stress pre-test. 

We then fit a linear model on this difference score and add a fixed intercept as we have no predictor. 

Consistent with hypothesis 1, stress was lower at post-test compared to pre-test across all participants. On average, participants decreased their stress score by 0.83 points ($b_0$ = 0.83, 95% CI = [0.72, 0.95], *p* < .001).

```{r}
evans_wide <- evans_wide %>% 
  mutate(stress_diff = stress_pre - stress_post)

lm_stress <- lm(stress_diff ~ 1, 
                data = evans_wide)

summary(lm_stress)

confint(lm_stress)
```
:::

### Hypothesis 2

10. In Hypothesis 2, we predict that the direct and indirect contact groups will have higher well-being and lower ill-being measures compared to the control group. For this question, we focus on social connectedness at post-test for a measure of well-being, so we expect higher scores in the direct and indirect groups. 

::: {.callout-tip}
#### Try this
Think about what techniques from Chapters 8 and 9 would let you test the hypothesis that the direct and indirect groups both increase social connectedness at post-test compared to the control group. Think of how you could focus on two groups at a time per analysis. 
:::

::: {.callout-note collapse="true"} 
#### Show me the task list

For the solution below, we apply two tests. We focus on the direct and control groups by filtering out indirect, then we focus on indirect and control by filtering out direct. 

To test the difference between two groups on one outcome, we need a model suitable for a between-subjects design. In Chapter 9, we covered [linear regression with one categorical predictor](#09-categorical) how to test the difference in an outcome between two groups. 

1. Fit a linear model on the outcome social connectedness post-test with a categorical predictor of group. You will need to fit two models, one where you filter out the indirect group and one where you filter out the direct group. 

2. Look at the intercept, what is the predicted value for the reference group? Look at the slope, what is the estimated change to the target group? For hypothesis 2, is the direct/indirect group higher on social connectedness compared to control? For hypothesis testing, is the *p*-value lower than alpha? 

3. What is the effect size? How much did the two groups differ on social connectedness? What is the confidence interval around the effect size? 

:::

::: {.callout-caution collapse="true"} 
#### Show me the solution

In the code below, we first fit a linear model on social connectedness post-test and focus on the direct and control groups. There are different ways you could ignore the indirect group, but here we filter the data as we pass the data to the `lm()` function.   

The direct group increased social connectedness post-test on average 0.15 points compared to the control group, but the difference was not statistically significant ($b_1$ = 0.15, 95% CI = [-0.09, 0.39], *p* = .207).

```{r}
lm_direct <- lm(social_post ~ GroupAssignment, 
                 data = filter(evans_wide,
                               GroupAssignment != "Indirect"))

summary(lm_direct)

confint(lm_direct)
```

The indirect group increased social connectedness post-test compared to the control group, but the difference was very small and not statistically significant ($b_1$ = 0.01, 95% CI = [-0.23, 0.25], *p* = .936).

```{r}
lm_indirect <- lm(social_post ~ GroupAssignment, 
                 data = filter(evans_wide,
                               GroupAssignment != "Direct"))

summary(lm_indirect)

confint(lm_indirect)
```

In @evans_repurposing_2023, the direct vs control comparison is significant, but they approached the analysis differently using a technique you will not learn until Research Methods 2. They control for pre-test scores by using it as an additional predictor, but you have not learnt about multiple regression yet. 
:::

### Hypothesis 3

11. Finally, the third hypothesis focuses on the difference between the two contact groups. They predict that the direct contact group will lead to higher well-being and lower ill-being compared to the indirect contact group. For this question, we focus on loneliness post-test for a measure of ill-being, so we expect lower scores in the direct group. 

::: {.callout-tip}
#### Try this
Think about what techniques from Chapters 8 and 9 would let you test the hypothesis that the direct group decreases loneliness at post-test compared to the indirect group.
:::

::: {.callout-note collapse="true"} 
#### Show me the task list

For this analysis, we focus on the final combination of groups, this time ignoring the control group.

To test the difference between two groups on one outcome, we need a model suitable for a between-subjects design. In Chapter 9, we covered [linear regression with one categorical predictor](#09-categorical) how to test the difference in an outcome between two groups. 

1. Fit a linear model on the outcome loneliness post-test with a categorical predictor of group. Filter the data to just focus on the direct and indirect contact groups. 

2. Look at the intercept, what is the predicted value for the reference group? Look at the slope, what is the estimated change to the target group? For hypothesis 3, is the direct group lower on loneliness compared to indirect? For hypothesis testing, is the *p*-value lower than alpha? 

3. What is the effect size? How much did the two groups differ on loneliness? What is the confidence interval around the effect size? 

:::

::: {.callout-caution collapse="true"} 
#### Show me the solution

In the code below, we first fit a linear model on loneliness post-test and focus on the direct and indirect groups. There are different ways you could ignore the indirect group, but here we filter the data as we pass the data to the `lm()` function.   

The direct group decreased loneliness post-test on average 0.14 points compared to the indirect group, but the difference was not statistically significant ($b_1$ = 0.14, 95% CI = [-0.01, 0.29], *p* = .060).

```{r}
lm_contact <- lm(lonely_post ~ GroupAssignment, 
                 data = filter(evans_wide,
                               GroupAssignment != "Control"))

summary(lm_contact)

confint(lm_contact)
```

Like hypothesis 2, @evans_repurposing_2023 report the direct vs indirect comparison as significant, but they approached the analysis differently using a technique you will not learn until Research Methods 2. They control for pre-test scores by using it as an additional predictor, but you have not learnt about multiple regression yet. 
:::

## Conclusion

Well done! Hopefully you recognised how far your skills have come to be able to do this independently, regardless of how many hints you needed. We are really starting to pile the skills up you have learnt so far, from simply using R Markdown files, to wrangling data, to visualising data, to finally applying modelling techniques for your inferential statistics. 

If you are curious, you can read @evans_repurposing_2023 to see how they walk through wrangling and analysing the data. Some of the techniques we do not cover in Research Methods 1, but they have some great features like highlighting common student mistakes. 
